---
title: "Calculating Hi-C coverage in each subgenome"
author: "Emma Sunde Tofastrud"
date: "2026-01-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Import libraries 

```{r}
library(ggplot2)
library(readr)
library(tidyverse)
library(knitr)
library(cowplot)
library(gridExtra)
library(gtools) # Used in plotting, using mixedsort which comes from this package the chromosome names were sorted numerically when plotting
```

# Calculate genome and subgenome sizes 

## *A. rouxii*

```{r}
# Import the chromosome sizes 
AmyRoux_sizes <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/Data/TADs/ggAmyRoux1_genome.fasta.sizes",col_names = c("contig", "size"))

# Sum the sizes of the chromosomes together to find the genome size
total_genome_AmyRoux <- sum(AmyRoux_sizes$size)

# Removes everything after the first underscore (_) and saves it in a new column named subgenome
AmyRoux_sizes$subgenome <- sub("_[^_]+.*", "", AmyRoux_sizes$contig) 

# Find the size of each subgenome
sizes_subgenome_AmyRoux <- aggregate(size ~ subgenome, data = AmyRoux_sizes, sum)

cat("Genome Size Summary (AmyRoux)\n")
cat("Total genome size: ", total_genome_AmyRoux, "\n\n", sep = "")
cat("Subgenome sizes:\n")
cat(
  paste0(
    sizes_subgenome_AmyRoux$subgenome,
    ": ",
    sizes_subgenome_AmyRoux$size,
    collapse = "\n"
  )
)
```

## *M. flavus*

```{r}
# Import the chromosome sizes 
MucFlav_sizes <- read_tsv("~/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/Contact maps/gzMucFlav1_genome.fasta.sizes",                           col_names = c("contig", "size"))

# Sum the sizes of the chromosomes together to find the genome size
total_genome_MucFlav <- sum(MucFlav_sizes$size) 

# Removes everything after the first underscore (_) and saves it in a new column named subgenome
MucFlav_sizes$subgenome <- sub("_[^_]+.*", "", MucFlav_sizes$contig) 

# Find the size of each subgenome

sizes_subgenome_MucFlav<- aggregate(size ~ subgenome, data = MucFlav_sizes, sum)

cat("Genome Size Summary (MucFlav)\n")
cat("Total genome size: ", total_genome_MucFlav, "\n\n", sep = "")
cat("Subgenome sizes:\n")
cat(
  paste0(
    sizes_subgenome_MucFlav$subgenome,
    ": ",
    sizes_subgenome_MucFlav$size,
    collapse = "\n"
  )
)
```

## *M. racemosus*

```{r}
# Import the chromosome sizes 
MucRace_sizes <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/Contact maps/gzMucRace1_genome.fasta.sizes",                           col_names = c("contig", "size"))

# Sum the sizes of the chromosomes together to find the genome size
total_genome_MucRace <- sum(MucRace_sizes$size) 

# Removes everything after the first underscore (_) and saves it in a new column named subgenome
MucRace_sizes$subgenome <- sub("_[^_]+.*", "", MucRace_sizes$contig) 

# Find the size of each subgenome

sizes_subgenome_MucRace<- aggregate(size ~ subgenome, data = MucRace_sizes, sum)

cat("Genome Size Summary (MucRace)\n")
cat("Total genome size: ", total_genome_MucRace, "\n\n", sep = "")
cat("Subgenome sizes:\n")
cat(
  paste0(
    sizes_subgenome_MucRace$subgenome,
    ": ",
    sizes_subgenome_MucRace$size,
    collapse = "\n"
  )
)
```

## Convert from bp to Mb

Divide on the genome and subgenome sizes on 1 000 000 to get per million base pairs instead of bp. 

```{r}
mb = 1000000
AmyRoux_genome_Mb = round(80325223/mb,2)
AmyRoux_sub1_Mb = round(40326961/mb,2)
AmyRoux_sub2_Mb = round(39998262/mb,2)

MucFlav_genome_Mb = round(75655998/mb,2)
MucFlav_sub1_Mb = round(42531630/mb,2)
MucFlav_sub2_Mb = round(33124368/mb,2)

MucRace_genome_Mb = round(121845080/mb,2)
MucRace_sub1_Mb = round(87150711/mb,2)
MucRace_sub2_Mb = round(34694369/mb,2)

# Table showing sizes in Mb for each species 
genome_table <- data.frame(
  Species = c("A. rouxii", "M. flavus", "M. racemosus"),
  Genome_Size_Mb = c(AmyRoux_genome_Mb, MucFlav_genome_Mb, MucRace_genome_Mb),
  Subgenome_1_Size_Mb = c(AmyRoux_sub1_Mb, MucFlav_sub1_Mb, MucRace_sub1_Mb),
  Subgenome_2_Size_Mb = c(AmyRoux_sub2_Mb, MucFlav_sub2_Mb, MucRace_sub2_Mb)
)
genome_table
```

# Calculate coverge estimate for genome and subgenomes using valid read pairs

To calculate the coverage for the whole genome, we used the following formula:

Total coverage: $Coverage_{total} = \frac{\text{Total number of valid pairs} \times 2 \times \text{Read length}}{\text{Genome size}}$.

To obtain a rough estimate of coverage for each subgenome, we used the following formula: 

Coverage per subgenome: $Coverage_{subgenome} = \frac{\text{# valid read pairs} \times 2 \times \text{Read length (bp)}}{\text{Subgenome size (bp)}}$

```{r}
# Read length for all species
read_length <- 146
```

## A. rouxii

```{r}
# The number of valid pairs per subgenome
total_valid_pairs_AR <- 9331600
valid_pairs_sub1_AR <- 3946555
valid_pairs_sub2_AR <- 3941782

# Size of the whole genome and subgenomes
total_genome_AmyRoux <- 80325223
sub1_size_AR <- 40326961
sub2_size_AR <- 39998262
```

```{r}
# Coverage for whole genome
coverage_genome_AR <- (total_valid_pairs_AR*2*read_length)/total_genome_AmyRoux
cat("Coverage for whole genome in A.rouxii: ", coverage_genome_AR, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 1
coverage_sub1_AR <- (valid_pairs_sub1_AR*2*read_length)/sub1_size_AR
cat("Coverage subgenome 1 in A.rouxii: ", coverage_sub1_AR, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 2
coverage_sub2_AR <- (valid_pairs_sub2_AR*2*read_length)/sub2_size_AR
cat("Coverage subgenome 2 in A.rouxii: ", coverage_sub2_AR, "\n\n", sep = "")
```

## M. flavus

```{r}
# The number of valid pairs per subgenome
total_valid_pairs_MF <- 16018096
valid_pairs_sub1_MF <- 24
valid_pairs_sub2_MF <- 16017819

# Size of the whole genome and subgenomes
total_genome_MucFlav <- 75655998
sub1_size_MF <- 42531630
sub2_size_MF <- 33124368
```

```{r}
# Coverage for whole genome
coverage_genome_MF <- (total_valid_pairs_MF*2*read_length)/total_genome_MucFlav
cat("Coverage for whole genome in M.flavus: ", coverage_genome_MF, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 1
coverage_sub1_MF <- (valid_pairs_sub1_MF*2*read_length)/sub1_size_MF
cat("Coverage subgenome 1 in M.flavus: ", coverage_sub1_MF, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 2
coverage_sub2_MF <- (valid_pairs_sub2_MF*2*read_length)/sub2_size_MF
cat("Coverage subgenome 2 in M.flavus: ", coverage_sub2_MF, "\n\n", sep = "")
```

## M. racemosus

```{r}
# The number of valid pairs per subgenome
total_valid_pairs_MR <- 997475
valid_pairs_sub1_MR <- 954937
valid_pairs_sub2_MR <- 42139

# Size of the whole genome and subgenomes
total_genome_MucRace <- 121845080
sub1_size_MR <- 87150711
sub2_size_MR <- 34694369
```

```{r}
# Coverage for whole genome
coverage_genome_MR <- (total_valid_pairs_MR*2*read_length)/total_genome_MucRace
cat("Coverage for whole genome in M.racemosus: ", coverage_genome_MR, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 1
coverage_sub1_MR <- (valid_pairs_sub1_MR*2*read_length)/sub1_size_MR
cat("Coverage subgenome 1 in M.racemosus: ", coverage_sub1_MR, "\n\n", sep = "")
```

```{r}
# Coverage for subgenome 2
coverage_sub2_MR <- (valid_pairs_sub2_MR*2*read_length)/sub2_size_MR
cat("Coverage subgenome 2 in M.racemosus: ", coverage_sub2_MR, "\n\n", sep = "")
```

# Calculate windowed coverage and estimate coverage per chromosome

## Import windowed files (20 kb windows for all species)

```{r}
# Importing coverage calculated for 20 kb windows in A. rouxii
coverage_AmyRoux <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/Data/Mapping coverage/HiC coverage for default HiC/ggAmyRoux1_coverage_20kb.txt",col_names = c("contig", "start", "end", "coverage"), skip = 0)
```

```{r}
# Importing coverage calculated for 20 kb windows in M. flavus
coverage_MucFlav<- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/gzMucFlav1_coverage_20kb.txt",col_names = c("contig", "start", "end", "coverage"), skip = 0)
```

```{r}
# Importing coverage calculated for 20 kb windows in M. racemosus
coverage_MucRace <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/gzMucRace1_coverage_20kb.txt",col_names = c("contig", "start", "end", "coverage"), skip = 0)
```

## Filtering 

### Filter out scaffolds from *A.rouxii*

```{r}
# Filter out all named scaffold, only keeping the ones named SUPER
coverage_AmyRoux <- coverage_AmyRoux %>%
  filter(
    !grepl("scaffold", contig, ignore.case = TRUE) & 
    !grepl("MT", contig, ignore.case = TRUE) # Filter out MT
  )
```

### Filter outsmall scaffolds in *M.flavus*

```{r}
# Start by merging the windowed coverage file with the chromosome sizes file, using contig as "merging factor". 
merged_MucFlav <- left_join(coverage_MucFlav, MucFlav_sizes, by = "contig")
```

```{r}
# Using the sizes column we filter out all contigs smaller than 100 000, removing scaffolds. 
filtered_MucFlav <- merged_MucFlav %>% filter(size >= 100000)
```

#### Filter out small scaffolds in *M.racemosus*

```{r}
# Start by merging the windowed coverage file with the chromosome sizes file, using contig as "merging factor". 
merged_MucRace <- left_join(coverage_MucRace, MucRace_sizes, by = "contig")
```

```{r}
# Filter out scaffolds smaller than 100 000.
filtered_MucRace1 <- merged_MucRace %>% filter(size >= 100000)
```

#### Filter out small scaffolds in *M.racemosus*

We do the same for M. racemosus as we did for *M.flavus*.

```{r}
merged_MucRace <- left_join(coverage_MucRace, MucRace_sizes, by = "contig")
```

```{r}
# Filter out scaffolds smaller than 100 000.
filtered_MucRace1 <- merged_MucRace %>% filter(size >= 100000)
```

# Mean coverage per chromosome 

```{r}
# Find mean coverage per chromosome by grouping by contig and then use summarise to find the mean coverage for that chromosome 
coverage_summary_AmyRoux <- coverage_AmyRoux  %>%
  group_by(contig) %>%
  summarise(mean_coverage = mean(coverage))
```

```{r}
coverage_summary_MucFlav <- filtered_MucFlav %>%
  group_by(contig) %>%
  summarise(mean_coverage = mean(coverage))
```

```{r}
coverage_summary_MucRace <- filtered_MucRace1 %>%
  group_by(contig) %>%
  summarise(mean_coverage = mean(coverage))
```

# Shorten chromosome names and add subgenome coloumn 

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
coverage_AmyRoux_clean <- coverage_summary_AmyRoux %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(contig_short = str_extract(contig, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
coverage_MucFlav_clean <- coverage_summary_MucFlav %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(contig_short = str_extract(contig, "[^_]+$")) 
```

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
coverage_MucRace_clean <- coverage_summary_MucRace %>% 
    # Create subgenome column based on first part before first "_"
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_"
  mutate(contig_short = str_extract(contig, "[^_]+$")) 
```

# Arrange chromosomes and subgenomes before plotting

```{r}
# Arrange so the numbers are shown by increasing number in all species
coverage_AmyRoux_clean <- coverage_AmyRoux_clean %>%
  # Sort by subgenome then chromosome
  arrange(subgenome, as.numeric(contig_short)) %>%   
  # unique for factor
  mutate(
    contig_unique = paste(subgenome, contig_short, sep = "_"),  
    contig_unique = factor(contig_unique, levels = contig_unique)
  )

coverage_MucFlav_clean <- coverage_MucFlav_clean %>%
  arrange(subgenome, as.numeric(contig_short)) %>%
  mutate(contig_short = factor(contig_short, levels = unique(contig_short)))

coverage_MucRace_clean <- coverage_MucRace_clean %>%
  arrange(subgenome, as.numeric(contig_short)) %>%
  mutate(contig_short = factor(contig_short, levels = unique(contig_short)))
```

# Plotting 

## Colour pallett for each species 

```{r}
species_palettes <- list(
  AmyRoux = c(
    "subgenome1" = "#0E7C7BFF",
    "subgenome2" = "#5FC0BF"
  ),
    MucFlav = c(
    "subgenome1" = "#B52626",
    "subgenome2" = "#E57373"
  ),
    MucRace = c(
    "subgenome1" = "#B25D91",
    "subgenome2" = "#cb91b4")
  )
```


```{r}
# Plot for A. rouxii 
AmyRoux_plot <- ggplot(coverage_AmyRoux_clean, 
                       aes(x = contig_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(title = "A.rouxii", x = "Chromosome", y = "Mean Coverage") +
  # show only chromosome number
  scale_x_discrete(labels = coverage_AmyRoux_clean$contig_short) + 
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

MucFlav_plot <- ggplot(coverage_MucFlav_clean, 
                       aes(x = contig_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") + # removes the legend showing subgenome
  labs(title = "M.flavus", x = "Chromosome", y = "Mean Coverage") +
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373")) + scale_x_discrete(expand = expansion(add = 0.9))


MucRace_plot <- ggplot(coverage_MucRace_clean, 
                       aes(x = contig_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none") + # removes the legend showing subgenome
  labs(title = "M. racemosus", x = "Chromosome", y = "Mean Coverage") +
  # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))

combined_plot_short <- plot_grid(
  AmyRoux_plot, MucFlav_plot, MucRace_plot,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

short_final_plot <- plot_grid( 
  combined_plot_short, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

short_final_plot
```

## Same y-axis

```{r}
# Plot for A. rouxii 
AmyRoux_plot_shared <- ggplot(coverage_AmyRoux_clean, 
                       aes(x = contig_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(title = "A.rouxii", x = "Chromosome", y = "Mean Coverage") +
  # show only chromosome number
  scale_x_discrete(labels = coverage_AmyRoux_clean$contig_short) + 
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))+ coord_cartesian(ylim = c(0, 300))

MucFlav_plot_shared <- ggplot(coverage_MucFlav_clean, 
                       aes(x = contig_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") + # removes the legend showing subgenome
  labs(title = "M.flavus", x = "Chromosome", y = "Mean Coverage") +
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373")) + scale_x_discrete(expand = expansion(add = 0.9)) + coord_cartesian(ylim = c(0, 300))


MucRace_plot_shared <- ggplot(coverage_MucRace_clean, 
                       aes(x = contig_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
        legend.position = "none") + # removes the legend showing subgenome
  labs(title = "M. racemosus", x = "Chromosome", y = "Mean Coverage") +
  # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4")) + coord_cartesian(ylim = c(0, 300))

combined_shared_axis <- plot_grid(
  AmyRoux_plot_shared, MucFlav_plot_shared, MucRace_plot_shared,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

shared_axis_HiCcoverageplot <- plot_grid( 
  combined_shared_axis, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

shared_axis_HiCcoverageplot
```


# Valid read pairs per megabase

## *A.rouxii*

```{r}
# To find the number of valid pairs per megabase we need to divide the number of valid pairs on the genome size (Mb).
total_valid_pairs_AR <- 9331600 

valid_pairs_AmyRoux_perMb <- total_valid_pairs_AR/AmyRoux_genome_Mb

cat("Valid pairs per Mb in A.rouxii:", valid_pairs_AmyRoux_perMb, "\n")
```

## *M.flavus*

```{r}
total_valid_pairs_MF <- 16018096 
valid_pairs_MucFlav_perMb <- total_valid_pairs_MF/MucFlav_genome_Mb

cat("Valid pairs per Mb in M.flavus:", valid_pairs_MucFlav_perMb, "\n")

```

## *M.racemosus*

```{r}
total_valid_pairs_MR <- 997475 
valid_pairs_MucRace_perMb <- total_valid_pairs_MR/MucRace_genome_Mb

cat("Valid pairs per Mb in M.racemosus", valid_pairs_MucRace_perMb, "\n")
```





