---
title: "PacBio and ONT coverage"
author: "Emma Sunde Tofastrud"
date: "2026-03-25"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Importing libraries 

```{r libraries}
library(ggplot2)
library(readr)
library(tidyverse)
library(cowplot)
library(gridExtra)
library(gtools) # Used in plotting, using mixedsort which comes from this package the contig names were sorted numarically when plotting
```

# Nanopore (ONT) (Only for A. rouxii)

## *A. rouxii*

```{r}
# Import file containing mean coverage for 20 kb sized windows
AR_ont <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/Data/Mapping coverage/ont_mean_coverage_20kb.txt", col_names = c("chrom", "start", "end", "coverage"))
```

### Filter out scaffolds

```{r}
# Filter out chromosomes with "scaffold" or "MT" in their name
AR_ont_filtered <- AR_ont[!grepl("Scaffold|MT", AR_ont$chrom), ]
```

### Finding the mean coverage for each chromosome 

```{r}
# Group by chromosome and find mean coverage for each chromosome
AR_ont_coverage <- AR_ont_filtered %>%
  group_by(chrom) %>%
  summarise(mean_coverage = mean(coverage)) # mean coverage
```

### Shorten chromosome names and add subgenome coloumn 

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
AR_ont_coverage_clean <- AR_ont_coverage %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(chrom, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(chrom_short = str_extract(chrom, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

### Arrange chromosomes and subgenomes before plotting

```{r}
# Arrange so the numbers are shown by increasing number in all species
AR_ont_coverage_clean <- AR_ont_coverage_clean %>%
  # Sort by subgenome then chromosome
  arrange(subgenome, as.numeric(chrom_short)) %>%   
  # unique for factor
  mutate(
    chrom_unique = paste(subgenome, chrom_short, sep = "_"),  
    chrom_unique = factor(chrom_unique, levels = chrom_unique)
  )
```

### Plotting the mean coverage for each chromosome

```{r}
AR_ont_coverage_barplot <- ggplot(AR_ont_coverage_clean, 
                       aes(x = chrom_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(
    title = "Average ONT Coverage per chromosome in A. rouxii (20kb windows)",
       x = "Chromosome", y = "Mean ONT Coverage") +
  # show only chromosome number
  scale_x_discrete(labels = AR_ont_coverage_clean$chrom_short) + 
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

AR_ont_coverage_barplot
```

# PacBio

## *A. rouxii*

```{r}
AR_pacbio <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/Data/Mapping coverage/ggAmyRoux1_pacbio_mean_coverage_20kb.txt", col_names = c("chrom", "start", "end", "coverage"))
```

### Filter out scaffolds

```{r}
# Filter out chromosomes with "scaffold" or "MT" in their name
AR_pacbio_filtered <- AR_pacbio[!grepl("Scaffold|MT", AR_pacbio$chrom), ]
```

### Finding the average coverage for each chromsome

```{r}
AR_pacbio_coverage <- AR_pacbio_filtered %>%
  group_by(chrom) %>%
  summarise(mean_coverage = mean(coverage))
```

### Shorten chromosome names and add subgenome coloumn 

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
AR_pacbio_coverage_clean <- AR_pacbio_coverage %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(chrom, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(chrom_short = str_extract(chrom, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

### Arrange chromosomes and subgenomes before plotting

```{r}
# Arrange so the numbers are shown by increasing number in all species
AR_pacbio_coverage_clean <- AR_pacbio_coverage_clean %>%
  # Sort by subgenome then chromosome
  arrange(subgenome, as.numeric(chrom_short)) %>%   
  # unique for factor
  mutate(
    chrom_unique = paste(subgenome, chrom_short, sep = "_"),  
    chrom_unique = factor(chrom_unique, levels = chrom_unique)
  )
```

### Plotting the mean coverage for each chromosome

```{r}
AR_pacbio_coverage_barplot <- ggplot(AR_pacbio_coverage_clean, 
                       aes(x = chrom_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(
    title = "Average PacBio Coverage per chromosome in A. rouxii (20kb windows)",
       x = "Chromosome", y = "Mean PacBio Coverage") +
  # show only chromosome number
  scale_x_discrete(labels = AR_ont_coverage_clean$chrom_short) + 
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

AR_pacbio_coverage_barplot
```

## *M. flavus*

```{r}
MF_pacbio <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucFlav1/gzMucFlav1_pacbio_mean_coverage_20kb_c.txt", col_names = c("chrom", "start", "end", "coverage"))
```

### Filter out scaffolds

```{r}
MF_sizes <- read_tsv("~/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/Contact maps/gzMucFlav1_genome.fasta.sizes",                           col_names = c("chrom", "size"))

merged_MF <- left_join(MF_pacbio, MF_sizes, by = "chrom")
# Filter out chromosomes smaller than 100 000
MF_pacbio_filtered <- merged_MF %>% filter(size >= 100000)
```

### Finding the average coverage for each chromsome

```{r}
MF_pacbio_coverage <- MF_pacbio_filtered %>%
  group_by(chrom) %>%
  summarise(mean_coverage = mean(coverage))
```

### Shorten chromosome names and add subgenome coloumn 

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
MF_pacbio_coverage_clean <- MF_pacbio_coverage %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(chrom, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(chrom_short = str_extract(chrom, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

### Arrange chromosomes and subgenomes before plotting

```{r}
# Arrange so the numbers are shown by increasing number in all species
MF_pacbio_coverage_clean <- MF_pacbio_coverage_clean %>%
  arrange(subgenome, as.numeric(chrom_short)) %>%
  mutate(chrom_short = factor(chrom_short, levels = unique(chrom_short))
  )
```

### Plotting the mean coverage for each chromosome

```{r}
MF_pacbio_coverage_barplot <- ggplot(MF_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(
    title = "Average PacBio Coverage per chromosome in M. flavus (20kb windows)",
       x = "Chromosome", y = "Mean PacBio Coverage") +
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373"))

MF_pacbio_coverage_barplot
```

## *M. racemosus*

```{r}
MR_pacbio <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/gzMucRace1_pacbio_mean_coverage_20kb.txt", col_names = c("chrom", "start", "end", "coverage"))
```

### Filter out scaffolds

```{r}
MR_sizes <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/Contact maps/gzMucRace1_genome.fasta.sizes",                           col_names = c("chrom", "size"))

merged_MR <- left_join(MR_pacbio, MR_sizes, by = "chrom")
# Filter out chromosomes smaller than 100 000
MR_pacbio_filtered <- merged_MR %>% filter(size >= 100000)
```

### Finding the average coverage for each chromsome

```{r}
MR_pacbio_coverage <- MR_pacbio_filtered %>%
  group_by(chrom) %>%
  summarise(mean_coverage = mean(coverage))
```

### Shorten chromosome names and add subgenome coloumn 

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
MR_pacbio_coverage_clean <- MR_pacbio_coverage %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(chrom, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(chrom_short = str_extract(chrom, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

### Arrange chromosomes and subgenomes before plotting

```{r}
# Arrange so the numbers are shown by increasing number in all species
MR_pacbio_coverage_clean <- MR_pacbio_coverage_clean %>%
  arrange(subgenome, as.numeric(chrom_short)) %>%
  mutate(chrom_short = factor(chrom_short, levels = unique(chrom_short))
  )
```

## Plotting the average coverage for each chromosome

```{r}
MR_pacbio_coverage_barplot <- ggplot(MR_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(
    title = "Average PacBio Coverage per chromosome in M. racemosus (20kb windows)",
       x = "Chromosome", y = "Mean PacBio Coverage") +
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))

MR_pacbio_coverage_barplot
```

# Combine plots 

```{r}
# Plot for A. rouxii 
AmyRoux_plot <- ggplot(AR_pacbio_coverage_clean, 
                       aes(x = chrom_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(title = "A.rouxii", x = "Chromosome", y = "Mean PacBio Coverage") +
  # show only chromosome number
  scale_x_discrete(labels = AR_pacbio_coverage_clean$chrom_short) + 
   # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

MucFlav_plot <- ggplot(MF_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") + # removes the legend showing subgenome
  labs(title = "M.flavus", x = "Chromosome", y = "Mean PacBio Coverage") +
   # Give each subgenome their own version of the species colour
   scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373"))


MucRace_plot <- ggplot(MR_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none")+ # removes the legend showing subgenome
  labs(title = "M. racemosus", x = "Chromosome", y = "Mean PacBio Coverage") +
  # Give each subgenome their own version of the species colour
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))


combined_plot_short <- plot_grid(
  AmyRoux_plot, MucFlav_plot, MucRace_plot,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

short_final_plot <- plot_grid( 
  combined_plot_short, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

short_final_plot
```

# Log-scaled 

```{r}
# Plot for A. rouxii 
AmyRoux_log <- ggplot(AR_pacbio_coverage_clean, 
                       aes(x = chrom_unique, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  scale_y_log10() +  # Logtransformed y-axis
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") +
  scale_y_log10(limits = c(1, 1000)) +
  labs(title = "A.rouxii", x = "Chromosome", y = "Mean PacBio Coverage\n(log scale)") +
  scale_x_discrete(labels = AR_pacbio_coverage_clean$chrom_short) + 
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

MucFlav_log <- ggplot(MF_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  scale_y_log10() +  # Logtransformed y-axis
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        legend.position = "none") +
  scale_y_log10(limits = c(1, 1000)) +
  labs(title = "M.flavus", x = "Chromosome", y = "Mean PacBio Coverage\n(log scale)") +
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373")) + 
  scale_x_discrete(expand = expansion(add = 0.9))


MucRace_log<- ggplot(MR_pacbio_coverage_clean, 
                       aes(x = chrom_short, y = mean_coverage, fill = subgenome)) +
  geom_bar(stat = "identity") +
  scale_y_log10() +  # Logtransformed y-axis
  theme_bw() +
  theme( axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none") +
  scale_y_log10(limits = c(1, 1000)) +
  labs(title = "M. racemosus", x = "Chromosome", y = "Mean PacBio Coverage\n(log scale)") + scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))

combined_logplot <- plot_grid(
  AmyRoux_log, MucFlav_log, MucRace_log,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

log_final_plot <- plot_grid( 
  combined_logplot, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

log_final_plot
```




