---
title: "Multi-mapping reads in A. rouxii, M. flavus and M. racemosus"
author: "Emma Sunde Tofastrud"
date: "2026-01-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(ggplot2)
library(readr)
library(tidyverse)
library(knitr)
library(cowplot)
library(gridExtra)
library(gtools) # Used in plotting, using mixedsort which comes from this package the contig names were sorted numerically when plotting
```

# Import, and merge R1 and R2 files

## *A. rouxii* 

```{r}
# Import file for R1
R1_AmyRoux <- read.table("/Users/emma/Documents/NMBU/Master/Master/Data/Mapping coverage/Multi-aligned default HiC/R1_mult_AmyRoux1.txt", header = FALSE)
colnames(R1_AmyRoux) <- c("contig", "start", "CIGAR", "align_score")

head(R1_AmyRoux)
```

```{r}
# Import file for R2
R2_AmyRoux <- read.table("/Users/emma/Documents/NMBU/Master/Master/Data/Mapping coverage/Multi-aligned default HiC/R2_mult_AmyRoux1.txt", header = FALSE)
colnames(R2_AmyRoux) <- c("contig", "start", "CIGAR", "align_score")

head(R2_AmyRoux)
```

```{r}
# Merge the files to show total number of multi-mappinged reads across the genome
AmyRoux_multi_mapping <- rbind(R1_AmyRoux, R2_AmyRoux)
```

## *M. flavus* 

```{r}
# Import file for R1
R1_MucFlav <- read.table("/Users/emma/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/Multi-aligned reads/R1_mult_MucFlav1.txt", header = FALSE)
colnames(R1_MucFlav) <- c("contig", "start", "CIGAR", "align_score")

head(R1_MucFlav)
```

```{r}
# Import file for R2
R2_MucFlav <- read.table("/Users/emma/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/Multi-aligned reads/R2_mult_MucFlav1.txt", header = FALSE)
colnames(R2_MucFlav) <- c("contig", "start", "CIGAR", "align_score")

head(R2_MucFlav)
```

```{r}
# Merge the files to show total number of multi-mapping reads across the genome
MucFlav_multi_mapping <- rbind(R1_MucFlav, R2_MucFlav)
```

## *M. racemosus* 

```{r}
# Import file for R1
R1_MucRace <- read.table("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/Multi-aligned reads/R1_mult_MucRace1.txt", header = FALSE)
colnames(R1_MucRace) <- c("contig", "start", "CIGAR", "align_score")

head(R1_MucRace)
```

```{r}
# Import file for R2
R2_MucRace <- read.table("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/Multi-aligned reads/R2_mult_MucRace1.txt", header = FALSE)
colnames(R2_MucRace) <- c("contig", "start", "CIGAR", "align_score")

head(R2_MucRace)
```

```{r}
# Merge the files to show total number of multi-mapping reads across the genome
MucRace_multi_mapping <- rbind(R1_MucRace, R2_MucRace)
```

# Multi-mapping reads per chromosome

## *A. rouxii*

```{r}
# Group by chromosome and find number of multi-mapping reads per chromosome
AmyRoux_reads_per_chrom <- AmyRoux_multi_mapping %>%
  group_by(contig) %>%
  summarise(count = n(), .groups = "drop")
```

```{r}
# Group by chromosome and find number of multi-mapping reads per chromosome
MucFlav_reads_per_chrom <- MucFlav_multi_mapping %>%
  group_by(contig) %>%
  summarise(count = n(), .groups = "drop")
```

```{r}
# Group by chromosome and find number of multi-mapping reads per chromosome
MucRace_reads_per_chrom <- MucRace_multi_mapping %>%
  group_by(contig) %>%
  summarise(count = n(), .groups = "drop")
```

# Filtering

## Filter out scaffolds and MT from *A. rouxii*

```{r}
# Filter out all named scaffold
AmyRoux_reads_per_chrom_filtered <- AmyRoux_reads_per_chrom %>%
  filter(
    !grepl("scaffold", contig, ignore.case = TRUE) & 
    !grepl("MT", contig, ignore.case = TRUE) # Filter out MT
  )
```

### Filter out small scaffolds in *M. flavus*

```{r}
# Import chromosome sizes 
MucFlav_sizes <- read_tsv("~/Documents/NMBU/Master/Master/gzMucFlav1/nf-core:HiC pipeline files/Contact maps/gzMucFlav1_genome.fasta.sizes",                           col_names = c("contig", "size"))
```

```{r}
# Join with multi-mapping file
merged_MucFlav <- left_join(MucFlav_reads_per_chrom, MucFlav_sizes, by = "contig")
```

```{r}
# Using the sizes column we filter out all contigs smaller than 100 000, removing scaffolds. 
MucFlav_reads_per_chrom_filtered <- merged_MucFlav %>% filter(size >= 100000)
```

### Filter out small scaffolds in *M. racemosus*

```{r}
# Import chromosome sizes 
MucRace_sizes <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/gzMucRace1/nfcore/Contact maps/gzMucRace1_genome.fasta.sizes",                           col_names = c("contig", "size"))
```

```{r}
# Join with multi-mapping file
merged_MucRace <- left_join(MucRace_reads_per_chrom, MucRace_sizes, by = "contig")
```

```{r}
# Using the sizes column we filter out all contigs smaller than 100 000, removing scaffolds. 
MucRace_reads_per_chrom_filtered <- merged_MucRace %>% filter(size >= 100000)
```

# Shorten chromosome names, divide into subgenomes and arrange

## *A. rouxii*

```{r}
# Shorten the chromosome name so it only includes chromosome number
# Add column indicating subgenome
AmyRoux_reads_per_chrom_clean <- AmyRoux_reads_per_chrom_filtered %>% 
  # Create subgenome column. Classify based on the part before the first "_". 
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_", only keeping the chromosome number
  mutate(contig_short = str_extract(contig, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

```{r}
# Arrange chromosomes for plotting
AmyRoux_reads_per_chrom_clean <- AmyRoux_reads_per_chrom_clean %>%
  # sort by subgenome then chromosome
  arrange(subgenome, as.numeric(contig_short)) %>%   
  # unique for factor
  mutate(
    contig_unique = paste(subgenome, contig_short, sep = "_"),  
    contig_unique = factor(contig_unique, levels = contig_unique)
  )
```

## *M. flavus*

```{r}
MucFlav_reads_per_chrom_clean <- MucFlav_reads_per_chrom_filtered %>% 
    # Create subgenome column based on first part before first "_"
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_"
  mutate(contig_short = str_extract(contig, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

```{r}
# Arrange chromosomes for plotting
MucFlav_reads_per_chrom_clean <-  MucFlav_reads_per_chrom_clean %>%
  arrange(subgenome, as.numeric(contig_short)) %>%
  mutate(contig_short = factor(contig_short, levels = unique(contig_short)))
```

## *M. racemosus*

```{r}
MucRace_reads_per_chrom_clean <- MucRace_reads_per_chrom_filtered %>% 
    # Create subgenome column based on first part before first "_"
  mutate(subgenome = str_extract(contig, "^[^_]+")) %>%
  # Remove everything before the last "_"
  mutate(contig_short = str_extract(contig, "[^_]+$")) %>% 
  mutate(subgenome = str_replace(subgenome, "subgenom(\\d+)", "subgenome\\1"))
```

```{r}
# Arrange chromosomes for plotting
MucRace_reads_per_chrom_clean <-  MucRace_reads_per_chrom_clean %>%
  arrange(subgenome, as.numeric(contig_short)) %>%
  mutate(contig_short = factor(contig_short, levels = unique(contig_short)))
```

# Multi-mappinged reads per Mb

```{r}
# Import AmyRoux sizes
AmyRoux_sizes <- read_tsv("/Users/emma/Documents/NMBU/Master/Master/Data/TADs/ggAmyRoux1_genome.fasta.sizes", col_names = c("contig", "size"))
```


```{r}
AmyRoux_reads_per_chrom_clean <- left_join(AmyRoux_reads_per_chrom_clean, AmyRoux_sizes, by = "contig") %>% mutate(mm_reads_per_Mb = count / (size / 1e6))
```

```{r}
MucFlav_reads_per_chrom_clean <- left_join(MucFlav_reads_per_chrom_clean, MucFlav_sizes, by = "contig")  

MucFlav_reads_per_chrom_clean <- MucFlav_reads_per_chrom_clean %>%  mutate(mm_reads_per_Mb = count / (size.x / 1e6))
```

```{r}
MucRace_reads_per_chrom_clean <- left_join(MucRace_reads_per_chrom_clean, MucRace_sizes, by = "contig")  

MucRace_reads_per_chrom_clean  <- MucRace_reads_per_chrom_clean  %>% mutate(mm_reads_per_Mb = count / (size.x / 1e6))
```

# Plotting 

```{r}
# Set upper limit for plots so that the y-axis is the same for all species
AmyRoux_Mb_plot <- ggplot(AmyRoux_reads_per_chrom_clean, 
                       aes(x = contig_unique, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "none") +
  labs(title = "A.rouxii", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = AmyRoux_reads_per_chrom_clean$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

MucFlav_Mb_plot <- ggplot(MucFlav_reads_per_chrom_clean, 
                       aes(x = contig_short, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "none") +
  labs(title = "M. flavus", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = MucFlav_reads_per_chrom_clean$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373"))

MucRace_Mb_plot <- ggplot(MucRace_reads_per_chrom_clean, 
                       aes(x = contig_short, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90), 
        legend.position = "none") +
  labs(title = "M. racemosus", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = MucRace_reads_per_chrom_clean$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))


combined_plot_short_Mb <- plot_grid(
  AmyRoux_Mb_plot, MucFlav_Mb_plot, MucRace_Mb_plot,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

short_final_plot_Mb <- plot_grid( 
  combined_plot_short_Mb, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

short_final_plot_Mb
```

# Multi-mapping reads after filtering out subgenomes with low coverage

```{r}
sub2_MucFlav <- MucFlav_reads_per_chrom_clean %>% filter(!grepl("^subgenome1", contig))
```

```{r}
sub1_MucRace <- MucRace_reads_per_chrom_clean %>% filter(!grepl("^subgenome2", contig))
```

```{r}
AmyRoux_Mb_plot <- ggplot(AmyRoux_reads_per_chrom_clean, 
                       aes(x = contig_unique, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0), 
        legend.position = "none") +
  labs(title = "A.rouxii", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = AmyRoux_reads_per_chrom_clean$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#0E7C7BFF", "subgenome2" = "#5FC0BF"))

MucFlav_Mb_plot_sub2 <- ggplot(sub2_MucFlav, 
                       aes(x = contig_short, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0), 
        legend.position = "none") +
  labs(title = "M. flavus", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = sub2_MucFlav$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#B52626", "subgenome2" = "#E57373"))

MucRace_Mb_plot_sub1 <- ggplot(sub1_MucRace, 
                       aes(x = contig_short, y = mm_reads_per_Mb, fill = subgenome)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0), 
        legend.position = "none") +
  labs(title = "M. racemosus", x = "Chromosome", y = "Multi-mapped reads/Mb") +
  scale_x_discrete(labels = sub1_MucRace$contig_short) + 
  scale_y_continuous(labels = scales::label_number(), limits = c(0, 1250000)) +
  scale_fill_manual(values = c("subgenome1" = "#B25D91", "subgenome2" = "#cb91b4"))


combined_plot_short_Mb <- plot_grid(
  AmyRoux_Mb_plot, MucFlav_Mb_plot_sub2, MucRace_Mb_plot_sub1,
  ncol = 1,
  labels = c("A", "B", "C"),
  label_size = 14,
  label_fontface = "bold",
  label_y = 1
)

short_final_plot_Mb <- plot_grid( 
  combined_plot_short_Mb, 
  ncol = 1, 
  rel_heights = c(0.1, 1)
)

short_final_plot_Mb

```





